---
title: "Spring Temperatures Above Zero Threshold"
author: "Annie Colgan"
date: "9/25/2020"
output: html_document
---
The Purpose of this script is to calculate the first day when smoothed air temperature crosses from below 0°C to above 0°C. The threshold and the code to calculate it were both adapted from Contosta et al. (2017) and Alix Contosta's Github repository. Running this code produces the following:

1. "above_zero_results.csv" - a csv file containing the results of trend analysis on the date of spring above zero threshold from 1935 - 2018 for both Pinkham Notch and Mt. Washington Summit.

2. "above_zero_for_graphs.csv" - a csv file containing the julian day of the temp above zero threshold for each year and each site joined with the results of trend analysis.

PLEASE NOTE: THIS CODE INVOLVES RUNNING 1000 ITERATIONS FOR EACH YEAR FOR EACH SITE AND CAN TAKE A WHILE TO RUN. YOU CAN CHANGE THE NUMBER OF ITERATIONS BY ADJUSTING THE "g" VARIABLE IN THE ABOVE ZERO THRESHOLD FUNCTION (line 88).

required input files: "pinkham_temp.csv", "summit_temp.csv"

Contosta, A. R., Adolph, A., Burchsted, D., Burakowski, E., Green, M., Guerra, D., ... & Routhier, M. (2017). A longer vernal window: the role of winter coldness and snowpack in driving spring transitions and lags. Global change biology, 23(4), 1610-1625.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(Kendall)
library(timetk)
library(trend)
library(tidyquant)
library(sweep)
library(forecast)
library(broom)
library(slider)
library(zoo)
library(BBmisc)#which.first
```

**Define Functions: **
The following code defines key functions that are used elsewhere in the script. 
```{r}
### WINTER JULIAN DAY FUNCTION ###
winter_doy_fun <- function(date){
  if(is.na(date == TRUE)){
    return(NA)
  }
  else{
  year <- year(date)
  month <- month(date)
  jday <- yday(date)
  winter.jday <- ifelse(leap_year(year) == TRUE, ifelse(month <=6, jday + 183, jday - 182), ifelse(month <= 6, jday + 184, jday -181))
}}

undo_winter_doy_fun <- function(wdoy, winter.year){
  if(is.na(wdoy == TRUE)){
    return(NA)
  }
  else{
    year <- ifelse(wdoy <= 184, winter.year-1, winter.year)
    doy <- ifelse(leap_year(year) == FALSE, ifelse(wdoy <= 184, wdoy + 181, wdoy - 184), ifelse(wdoy <= 185, wdoy + 181, wdoy -184))
    origin.date <- lubridate::ymd(year, truncated = 2)
    date <- as.Date(doy, origin = as.character(origin.date-1))
    return(date)
  }
}

### Vernal Window Functions ###
##adapted by Annie Colgan from code posted on github by Alix Contosta, https://github.com/Contosta/vernal-windows/blob/master/air_temp_zero.R

#Mode function
#this function is called within the above zero threshold to identify to mode of date. 
#input: a vector containing values of type num.
#output: a single value corresponding to the mode of the input vector. 
Mode <- function(x, na.rm = FALSE) {
  if(na.rm){
    x = x[!is.na(x)]
  }
  
  ux <- unique(x)
  return(ux[which.max(tabulate(match(x, ux)))])
}

## Above zero threshold function ##
#This function uses a monte carlo approach to identify the threshold at which the smoothed average daily temperature first gets above zero. 

#input: a dataframe containing a date column and a average temperature column (°C) with data for a single winter year (July 1st through June 30th). 

#output: a single value corresponding to the first winter day of year when the smoothed average daily temperature is above zero. 

above.zero.fun <- function(data){
# User defined variables
g <- 1000#number of iterations to run
break.est.var <- 15 #the number of days to vary the search window size by

#initialize vector
output.doys <- vector(length = g)

#create timeseries
temp_date <- data$date
z1 <- zoo(data$average, temp_date)
z <- z1[!is.na(z1)]

#subset timeseries
year <- year(temp_date[200])
start2 <- ymd(paste (toString(year-1), "-11-01", sep = "", collapse = NULL))
end2 <- ymd(paste (toString(year), "-06-30", sep = "", collapse = NULL))
time.series <- window(z, start = start2, end = end2)


#Looping through the iterations
for(p in 1:g){
  w1 <- sample(seq(from = 5, to = 80), size = 1)
  w <- floor( w1 / 2) * 2 + 1 #converts numbers to odd integers
  z.median <- rollmedian(time.series, k = w)
  
  #finding start day through randomly generated variable
  startseed <- ymd(paste (toString(year), "-01-15", sep = "", collapse = NULL)) 
  endseed <- ymd(paste (toString(year), "-07-15", sep = "", collapse = NULL))
  var <- break.est.var
  start.day <- sample(seq( from = startseed - var,
                         to = startseed + var, by = "day"), size = 1 )
  end.day <- sample(seq( from = endseed - var, 
                          to = endseed + var, by = "day"), size = 1 )
  
  if(end.day-start.day < 10) {next}
  
  #running piecewise linear regression
  med.for.regression.temp <- window(z.median, start = start.day, end = end.day)
  temp.series <- coredata(med.for.regression.temp)
  date.series <- index(med.for.regression.temp)
  lengz <- length(med.for.regression.temp)
  #run through the dataset starting at the end and find the 0 threshold
  if (min(temp.series)>0) {next}
  
  for(day in 1:lengz)
  {
    if(temp.series[(lengz-day)]<0)
    {begin.temp <- winter_doy_fun(date.series[(lengz-day)])}
    if(temp.series[(lengz-day)]<0)
    {break}
  }
  
  output.doys[p] <- begin.temp
  }

above.zero.threshold <- Mode(output.doys)
}


### TIDY SENS SLOPE FUNCTION ###
#tidy() function doesn't work for sens slope. This function works instead. 
#input: the output of the sens function
#output: the sens estimate, sens statistic, and sens.p.value, unlisted
tidy_sens_fun <- function(sens){
  tidy_sens <- tibble(sens.slope.estimate = sens["estimates"] %>% unlist, 
                            sens.slope.statistic = sens['statistic'] %>% unlist,
                      sens.p.value = sens["p.value"]%>%unlist)
}


```



**1. Load temperature and snow data and join: **
The following code loads the data from csv files and joins them into a single table. 
```{r}
###TEMPERATURE DATA###

#Loads Pinkham temperature data
#Starts Jan 1, 1935
PN_temp_df <- read_csv("pinkham_temp.csv")%>%
  mutate(date = paste(year, month, day, sep = ' ')%>% #adds date column
  ymd(),
  site = 'pinkham')%>% #adds site column
  rename(maxC = MaxT_PN, minC = MinT_PN, average = temp_c) #renames columns to match summit data
  
#Loads Summit temperature data
#starts Jan 1 1935
S_temp_df <- read_csv('summit_temp.csv')%>%
  rename(month = mo)%>% #renames month column to match pinkham data
  mutate(date = ymd(paste(year, month, day, sep = "-")), #adds date column
         site = 'summit')%>% #adds site column
  select(-jday) 


#joins pinkham and summit
temp_df <- PN_temp_df %>%
  bind_rows(S_temp_df)%>%
   mutate(winter.year = ifelse(month <= 6, year, year+1))

temp_df
```


```{r}
above.zero.threshold <- 20#defines how many days are allowed to be missing data in a given year. If the number of days without data is greater than the "above.zero.threshold" that year is excluded from the analysis. 

missing_data <- temp_df %>%
  group_by(site, winter.year)%>%
  mutate(avg.missing = ifelse(is.na(average)==TRUE, 1, 0))%>%
  summarize(avg.missing.sum = sum(avg.missing))

above_zero_df <- temp_df %>%
  filter(winter.year > 1935 & winter.year < 2018) %>%
  select(site, winter.year, average, date) %>%
  group_by(site, winter.year)%>%
  nest()%>%
  left_join(missing_data, by = c("site", "winter.year"))%>%
  mutate(above.zero.threshold =  ifelse(winter.year < 1936, NA, ifelse(avg.missing.sum > above.zero.threshold, NA,
                                    as.numeric(map(.x = data,
                                .f = above.zero.fun)))))
```
4. Calculate Sens Slope and Mann Kendall 
```{r}
#nests years for each indice
above_zero_nested <- above_zero_df %>%
  ungroup()%>%
  mutate(date = ymd(winter.year, truncated = 2))%>%
  select(site, date, above.zero.threshold)%>%
  filter(is.na(above.zero.threshold)==FALSE)%>%
  mutate(above.zero.threshold = as.numeric(above.zero.threshold))%>%
  group_by(site)%>% 
  nest()

#add column containing time series
above_zero_ts <- above_zero_nested %>%
  mutate(data.ts = map(.x = data,
                       .f = tk_ts,
                       select = -date,
                       start = 1936,
                       freq = 1))

#create summary df with results of mann kendall and sens slope tests
above_zero_summary <- above_zero_ts %>%
  mutate(mk = map(data.ts, MannKendall),
         sens = map(data.ts, sens.slope))%>%
  mutate(tidy_mk = map(mk, tidy),
         tidy_sens = map(sens, tidy_sens_fun))%>%
  unnest(c(tidy_sens, tidy_mk))



```


**6. Combine metric data with results of trend analysis for graphing**

```{r}
#create neat data table for joining
above_zero_for_join <- above_zero_summary %>%
  mutate(trend = ifelse(p.value >= 0.05, 'no trend', ifelse(kendall_score > 0, "increasing", "decreasing")),
         slope = sens.slope.estimate,
         days_per_decade = 10*abs(slope))%>%
  select(site, trend, slope, days_per_decade, sens.p.value)

above_zero_for_join%>%
  write_csv("above_zero_results.csv")

above_zero_for_join <- read_csv("above_zero_results.csv")

#join
above_zero_for_graphs <- above_zero_df %>%
  mutate(above.zero.threshold = as.numeric(above.zero.threshold))%>%
  left_join(above_zero_for_join, by = 'site')%>%
  mutate(date = map(.f = undo_winter_doy_fun,
                    .x = above.zero.threshold,
                    winter.year = winter.year),
         date = as_date(as.numeric(date)),
         date_for_graph = ymd(format(date, "1936-%m-%d")))%>%
  select(-data)%>%
  write_csv("above_zero_for_graphs.csv")

```



