---
title: "CV"
author: "Annie Colgan"
date: "9/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(ggpubr)
```

You can ignore these top parts, it's just the code for loading in the data. 
```{r}
#Loads Pinkham temperature data
PN_temp_df <- read_csv("pinkham_temp.csv")%>%
  mutate(date = paste(year, month, day, sep = ' ')%>% #adds date column
  ymd(),
  site = 'pinkham')%>% #adds site column
  rename(maxC = MaxT_PN, minC = MinT_PN, average = temp_c) #renames columns to match summit data
  
#Loads Summit temperature data
S_temp_df <- read_csv('summit_temp.csv')%>%
  rename(month = mo)%>% #renames month column to match pinkham data
  mutate(date = ymd(paste(year, month, day, sep = "-")), #adds date column
         site = 'summit')%>% #adds site column
  select(-jday) 

#Joins pinkham and summit data into single data frame
temp_df <- PN_temp_df %>%
  bind_rows(S_temp_df) 

temp_df
```


**CV starts here:**
I calculated the CV for each day of the year just to see what would happen...
```{r}
#CV of mean temperature for each month 
cv_df <- temp_df %>%
  mutate(doy = yday(date))%>%
  group_by(doy, site)%>%
  summarize(mean = mean(average, na.rm = TRUE),
            sd = sd(average, na.rm = TRUE),
            cv = abs((sd/mean)*100)) #average is the average daily temperature

cv_df
```
Here you can see that in some cases standard deviation is almost as large as the mean, which is part of why we're getting such high CV values. CV is reported either as a ratio of the sd to the mean (a value between 0 and 1) or as the percent sd is of the mean. I've multiplied by a hundred so these are percents. 

```{r}
cv_for_graph <-  cv_df %>%
  mutate(site = ifelse(site == 'summit', 'Summit', 'Pinkham'),
         date = as.Date(doy, origin = "1931-12-31"))

#daily 
doy_o <- cv_for_graph %>%
  ggplot(aes(x = date, y = cv, shape = site, color = site)) +
  geom_point(size = 1)+
  theme_bw()+
  scale_x_date(date_labels = "%B")+
  labs(y = "Coefficient of Variance (%)", x = "Day of the Year" )+
  scale_color_manual(values = c("grey1", "grey45"))+
  theme(legend.position = c(0.13,0.8),
        legend.title = element_blank(),
        text = element_text(family = "Helvetica", size = 10),
        axis.text = element_text(size = 10),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank())

ggsave('cv_doy_original.png', width = 5.25, height = 3.28)


#log scale on the y axis
doy_log <- cv_for_graph %>%
  ggplot(aes(x = date, y = cv, shape = site, color = site)) +
  geom_point(size = 1)+
  theme_bw()+
  scale_x_date(date_labels = "%b-%d")+
  scale_y_continuous(trans = 'log10')+
  labs(y ="Coefficient of Variance (%)", x = "Day of the Year" )+
  scale_color_manual(values = c("grey1", "grey45"))+
  theme(legend.position = c(0.13,0.8),
        legend.title = element_blank(),
        legend.text = element_text(family = "Helvetica", size = 8),
        text = element_text(family = "Helvetica", size = 10))

  


#zooming in so that we can see most of the data better...
doy_z <- cv_for_graph %>%
  ggplot(aes(x = date, y = cv, shape = site, color = site)) +
  geom_point(size = 1)+
  theme_bw()+
  scale_x_date(date_labels = "%b-%d")+
  labs(y ="Coefficient of Variance (%)", x = "Day of the Year" )+
  scale_color_manual(values = c("grey1", "grey45"))+
  theme(legend.position = c(0.13,0.8),
        legend.title = element_blank(),
        text = element_text(family = "Helvetica", size = 10))+
  ylim(-100, 500)

ggsave("cv_doy_zoomed.png", width = 5.25, height = 3.28)

ggarrange(doy_o, doy_z, nrow =2, align = "h")

ggsave("cv_doy_both.png", width = 5.25, height = 3.28)


#CV of mean temperature for each decade
cv_decade_df <- temp_df %>%
  mutate(decade = floor(year/10)*10,
         site = ifelse(site == 'summit', 'Summit', 'Pinkham'))%>% #adds decade column
  group_by(decade, site)%>% #groups by decade and site (summit and pinkham)
  summarize(cv = abs((sd(average, na.rm=TRUE)/mean(average, na.rm = TRUE))*100)) #calculates cv


#visualize
cv_decade <- cv_decade_df %>%
  filter(decade > 1930 & decade < 2010)%>%
  ggplot(aes(x = decade, y = cv, color = site, shape = site)) +
  geom_point()+
  theme_bw()+
  labs(y = "Coefficient of Variance (%)", x = "Decade (1940s-2000s)" )+
  scale_color_manual(values = c("grey1", "grey45"))+
  theme(legend.position = "none",
        legend.title = element_blank(),
        text = element_text(family = "Helvetica", size = 10))
  
ggarrange (doy_log, cv_decade, nrow = 2, labels = c("a", "b"))
ggsave("cv_doylog_dec.jpeg", width = 5.25, height = 6.74, units = "in", dpi = 1000)

```

Each dot represents the cv of a single day of the year. The two graphs are the same, I just zoomed in in the second one so we could see better. Color represents site (summit vs pinkham). Just like Georgia was hypothesizing, the summit is more variable in summer, though interestingly it's less variable in winter. Something to do with the PBL? And look at those peaks at the start and end of the growing season!

The rest of this is just the same graphs we saw yesterday except I took absolute values to make them easier to compare. Let me know if there is anything else either of you would like to see!

```{r}
#CV doy change over time 30 year bins
cv_df <- temp_df %>%
  mutate(doy = yday(date),
         period = ifelse(year(date)<1977, 'early', 'late'))%>%
  group_by(doy, site, period)%>%
  summarize(mean = mean(average, na.rm = TRUE),
            sd = sd(average, na.rm = TRUE),
            cv = abs((sd/mean)*100))

cv_df %>% ggplot(aes(x = doy, y = cv, color = site))+
  geom_point()+
  facet_wrap(~period)+
  theme_bw()

#zoomed in
cv_df %>% ggplot(aes(x = doy, y = cv, color = site))+
  geom_point()+
  facet_wrap(~period)+
  theme_bw()+
  ylim(-100, 500)

```


```{r}
#CV doy change over time, decades
cv_df <- temp_df %>%
  mutate(doy = yday(date),
         decade = floor(year/10)*10)%>%
  group_by(doy, site, decade)%>%
  summarize(mean = mean(average, na.rm = TRUE),
            sd = sd(average, na.rm = TRUE),
            cv = abs((sd/mean)*100))

cv_df %>%
  ggplot(aes(x = doy, y = cv, color = site))+
  geom_point(size = .1)+
  facet_wrap(~decade)+
  theme_bw()

#zoom
cv_df %>%
  ggplot(aes(x = doy, y = cv, color = site))+
  geom_point(size = .1)+
  facet_wrap(~decade)+
  theme_bw()+
  ylim(-100, 500)#zoomed

#another way to display
cv_df %>%
  ggplot(aes(x = doy, y = cv, color = decade))+
  geom_point(size = .1)+
  facet_wrap(~site)+
  theme_bw()+
  ylim(-100, 500)#zoomed
```


```{r}
#CV of mean temperature for each month 
cv_df <- temp_df %>%
  group_by(month, site)%>%
  summarize(mean = mean(average, na.rm = TRUE),
            sd = sd(average, na.rm = TRUE),
            cv = abs((sd/mean)*100))%>% #average is the average daily temperature
  mutate(month = month(month, label = TRUE))

cv_df

#visualize
cv_df %>%
  ggplot(aes(x = month, y = cv, color = site)) +
  geom_point()+
  theme_bw()

ggsave('cv_month.png')

```


```{r}
#CV of mean temperature for each decade
cv_decade_df <- temp_df %>%
  mutate(decade = floor(year/10)*10)%>% #adds decade column
  group_by(decade, site)%>% #groups by decade and site (summit and pinkham)
  summarize(cv = abs((sd(average, na.rm=TRUE)/mean(average, na.rm = TRUE))*100)) #calculates cv

cv_decade_df

#visualize
cv_decade_df %>%
  ggplot(aes(x = decade, y = cv, color = site)) +
  geom_point()+
  theme_bw()

```



