---
title: "Trends in Climate Indicators"
author: "Annie Colgan"
date: "9/23/2020"
output: html_document
---
The purpose of this code is to calculate climate indicators and analyze trends in those indicators for Pinkham Notch and the Summit of Mount Washington. 

1. "climate_indices_trend_results.csv" - a csv file containing results of the analysis

2. "growing.season.end.png" - a graph showing data on the end of the growing season at the summit of mount washington over time. 

required input files: "pinkham_temp.csv", 'summit_temp.csv', 'PN_snow.csv', 'summit_snow.csv'


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(Kendall)
library(timetk)
library(trend)
library(tidyquant)
library(sweep)
library(forecast)
library(broom)
library(slider)
library(BBmisc)#which.first

```

**Define Functions: **
The following code defines key functions that are used elsewhere in the script. 
```{r}
### WINTER JULIAN DAY FUNCTION ###
winter_doy_fun <- function(date){
  if(is.na(date == TRUE)){
    return(NA)
  }
  else{
  year <- year(date)
  month <- month(date)
  jday <- yday(date)
  winter.jday <- ifelse(leap_year(year) == TRUE, ifelse(month <=6, jday + 183, jday - 182), ifelse(month <= 6, jday + 184, jday -181))
  }}

### TIDY SENS SLOPE FUNCTION ###
#tidy() function doesn't work for sens slope. This function works instead. 
#input: the output of the sens function
#output: the sens estimate, sens statistic, and sens.p.value, unlisted
tidy_sens_fun <- function(sens){
  tidy_sens <- tibble(sens.slope.estimate = sens["estimates"] %>% unlist, 
                            sens.slope.statistic = sens['statistic'] %>% unlist,
                      sens.p.value = sens["p.value"]%>%unlist)
}

```

**1. Load temperature and snow data and join: **
The following code loads the data from csv files and joins them into a single table. 
```{r}
###TEMPERATURE DATA###

#Loads Pinkham temperature data
#Starts Jan 1, 1935
PN_temp_df <- read_csv("pinkham_temp.csv")%>%
  mutate(date = paste(year, month, day, sep = ' ')%>% #adds date column
  ymd(),
  site = 'pinkham')%>% #adds site column
  rename(maxC = MaxT_PN, minC = MinT_PN, average = temp_c) #renames columns to match summit data
  
#Loads Summit temperature data
#starts Jan 1 1935
S_temp_df <- read_csv('summit_temp.csv')%>%
  rename(month = mo)%>% #renames month column to match pinkham data
  mutate(date = ymd(paste(year, month, day, sep = "-")), #adds date column
         site = 'summit')%>% #adds site column
  select(-jday) 

###SNOW DATA###

#Load Pinkham snow data
#stars Jan 1, 1930
PN_snow_df <- read_csv('PN_snow.csv')%>%
  select(DATE, `SNOW(MM)`, `SNAW (MM)`)%>%
  rename(date = DATE, snow.fall = `SNOW(MM)`, snow.depth = `SNAW (MM)`)%>%#renames columns, fills missing snow depth value with previous
  mutate(date = mdy(date),#converts date to type date
         site = 'pinkham')


#Load Summit snow data
#starts Jan 1, 1948
S_snow_df <- read_csv('summit_snow.csv')%>%
  mutate(date = mdy(Date),#adds date column with data in type date
         Snowfall = as.numeric(ifelse(Snowfall == "M", NA, ifelse(Snowfall == "T", (0), Snowfall))), #replaces M with NA and T with 0, to be consistent with PN
         SnowDepth = as.numeric(ifelse(SnowDepth == "M", NA, ifelse(SnowDepth == "T", (.5), Snowfall))),#replaces T with .5, half the 1" threshold for snow depth
         snow.fall = Snowfall*25.4, #converts to mm
         snow.depth = SnowDepth*25.4, #converts to mm
         site = 'summit')

###JOIN DATA###
#join pinkham
#starts 1930
PN_df <- PN_snow_df %>%
  full_join(PN_temp_df)

#join summit
#start 1935
S_df <- S_temp_df %>%
  full_join(S_snow_df)

#joins pinkham and summit
climate_df <- PN_df %>%
  bind_rows(S_df)

climate_df
```



**2. Create a new data frame containing index values**

```{r}
indices_df <- climate_df %>%
  mutate(year = year(date),
         month = month(date),
         day = day (date),
        winter.year = ifelse(month <= 6, year, year+1), #add winter.year column corresponding to the year for Jan-Jun
         winter = ifelse(month >= 11 | month <= 5, TRUE, FALSE), #defines winter as November through May
         jday = yday(date), #adds julian day column
         winter.jday = winter_doy_fun(date), #adds winter julian day column
        
         #the following adds column containing 1,0 for each index
         frost.day = ifelse(minC < 0, 1, 0),
         ice.day = ifelse(maxC < 0, 1, 0),
         snowmaking.day = ifelse(minC < -5 & month >= 9 & jday < 359, 1, 0),
         spruce.day = ifelse(minC < -32, 1, 0),
         winter.above.zero.day = ifelse(winter == TRUE & maxC > 0, 1, 0),
         thaw.day = ifelse(winter == TRUE & minC > 0, 1, 0),
         lag1 = lag(thaw.day, order_by = date),
         lag2 = lag(lag1, order_by = date),
         lag3 = lag(lag2, order_by = date),
         lead1 = lead(thaw.day, order_by = date),
         lead2 = lead(lead1, order_by = date),
         lead3 = lead(lead2, order_by = date),
         thaw.event.day = ifelse(thaw.day == 0, 0, ifelse(lag3+lag2+lag1+thaw.day == 4 | lag2 +lag1 +thaw.day + lead1 == 4 | lag1 + thaw.day + lead1 + lead2 == 4 | thaw.day + lead1 + lead2 + lead3 == 4, 1, 0)),
         thaw.event.day = ifelse(winter == TRUE, thaw.event.day, 0),
         frozen.ground.day = ifelse(maxC < 0 & snow.depth == 0, 1, 0),
         mud.day = ifelse(maxC > 0 & snow.depth == 0 & month < 6 | month > 10, 1, 0),
         snow.covered = ifelse(snow.depth > 0, 1, 0),
         snow.day = ifelse(snow.fall > 0, winter.jday, NA),
        growing.start.day = ifelse(average > 5, 1, 0),
         growing.start.days = ifelse(slide(growing.start.day, ~mean(.x), .before = 5) == 1, TRUE, FALSE), #last day of first day period
         growing.end.day = ifelse(month > 6 & average < 5, 1, 0),
          growing.end.days = ifelse(slide(growing.end.day, ~mean(.x), .after = 5) == 1, TRUE, FALSE ), #first day of 6 day period
        freeze.day.24.spring = ifelse(minC < -4.4 & month < 7, TRUE, FALSE),
        freeze.day.24.fall = ifelse(minC < -4.4 & month >= 7, TRUE, FALSE),
         max.missing = ifelse(is.na(maxC)==TRUE, 1, 0),
         min.missing = ifelse(is.na(minC)==TRUE, 1, 0),
         fall.missing = ifelse(is.na(snow.fall) == TRUE, 1, 0),
        depth.missing = ifelse(is.na(snow.depth)== TRUE, 1, 0),
         average.missing = ifelse(is.na(average)==TRUE, 1, 0))

indices_df
      
```

**3. Summarize indices by year **
```{r}
### DEFINE THRESHOLDS FOR DEALING WITH MISSING DATA ###
threshold <- 10 #defines the number of days of data allowed to be missing each year. If the number of missing days exceeds this threshold, the value for that year is 'NA'.

snow.threshold <- 10#defines the number of days of snow data allowed to be missing each snow season. Snow season is defined as the period between November 1st and May 31st. 
winter.start <- 123 #November 1st as wdoy
winter.end <- 335 #May 31 as wdoy


#Joins first and last snow information with indices_df and replaces
indices_df_2 <- indices_df %>% 
  mutate(depth.missing = ifelse (winter.jday > winter.start  & winter.jday < winter.end & depth.missing == 1, 1, 0), #adjusts depth missing to only care about days between first and last snow fall
         fall.missing = ifelse(winter.jday > winter.start & winter.jday < winter.end & fall.missing == 1, 1, 0))

### SUMMARIZE INDICES THAT ARE COUNTED BY WINTER YEAR ###
winter_indices <- indices_df_2 %>%
  group_by(winter.year, site) %>%
  summarize(
    frost.days = ifelse(sum(min.missing) < threshold, sum(frost.day, na.rm = TRUE), NA), 
    ice.days = ifelse(sum(max.missing)<threshold, sum(ice.day, na.rm = TRUE), NA), 
    spruce.days = ifelse(sum(min.missing)<threshold, sum(spruce.day, na.rm = TRUE), NA),
    winter.above.zero.days = ifelse(sum(max.missing)<threshold, sum(winter.above.zero.day, na.rm = TRUE), NA),
  thaw.event.days = ifelse(sum(min.missing)<threshold, sum(thaw.event.day, na.rm = TRUE), NA),
    snow.covered.days = ifelse(sum(depth.missing) < snow.threshold, sum(snow.covered, na.rm = TRUE), NA),
  frozen.ground.days = ifelse(sum(depth.missing) < snow.threshold & sum(max.missing) < threshold, sum(frozen.ground.day, na.rm = TRUE), NA),
  mud.days = ifelse(sum(depth.missing) < snow.threshold & sum(max.missing) < threshold, sum(mud.day, na.rm = TRUE), NA),
  snowmaking.days = ifelse(sum(min.missing)<threshold, sum(snowmaking.day, na.rm = TRUE), NA),)%>%
  mutate(year = winter.year,
         date = ymd(year, truncated = 2))

### SUMMARIZE INDICES THAT ARE COUNTED BY CALENDAR YEAR ###
summer_indices <- indices_df %>%
  group_by(year, site)%>%
  summarize(
    growing.season.start = ifelse(sum(average.missing) < threshold,
                                  jday[which.first(growing.start.days)], NA),
    growing.season.end = ifelse(sum(average.missing) < threshold,
                                jday[which.first(growing.end.days)], NA),
    spring.24.day = ifelse(sum(min.missing) < threshold,
                        jday[which.last(freeze.day.24.spring)], NA),
    fall.24.day = ifelse(sum(min.missing) < threshold,
                    jday[which.first(freeze.day.24.fall)], NA))%>%
  mutate(growing.days = growing.season.end - growing.season.start,
         growing.season.24 = as.numeric(fall.24.day) - as.numeric(spring.24.day))



### JOIN ALL INDICES INTO ONE DATA FRAME ###
indices_by_year <- left_join(winter_indices, summer_indices, by = c("year", "site"))%>%
  relocate(growing.season.start:growing.season.24, .before = snowmaking.days)%>%
  filter(winter.year > 1935 & winter.year <= 2018)%>%
  ungroup()%>%
  select(-year, -winter.year)%>%
  pivot_longer(cols = frost.days:snowmaking.days, names_to = 'metric', values_to = 'value')'
  
  

```
Compute the n of days between the start and end of freezing season at the beginning and end of timeseries for each site – would be good to say the freezing season was X days shorter by the 2010s as compared to 1930s, as well as compare for the 2 sites
```{r}
growing_season_length <- indices_by_year %>% 
  filter(metric == "spring.24.day" | metric == "fall.24.day")%>%
  mutate(year = year(date))%>%
  select(-date)%>%
  pivot_wider(names_from = c(metric), values_from = value)%>%
  mutate(length = fall.24.day - spring.24.day,
         period = ifelse(year < 1946, "1936-1945", ifelse(year > 2007, "2008-2018", "other")))%>%
  group_by(site, period)%>%
  summarize(average_length = mean(length, na.rm = TRUE),
            sd_length = sd(length, na.rm = TRUE))%>%
  write_csv("growing_season_length_comparison.csv")
 
  



```

4. Calculate Sens Slope and Mann Kendall 
```{r}
#nests years for each indice
indices_nested <- indices_by_year %>%
  filter(is.na(value)==FALSE)%>%
  mutate(value = as.numeric(value))%>%
  group_by(metric, site)%>% 
  nest()

#add column containing time series
indices_nested_ts <- indices_nested %>%## Switch this back to tk_ts
  mutate(data.ts = map(.x = data,
                       .f = tk_ts,
                       select = -date,
                       start = 1936,
                       freq = 1))

#create summary df with results of mann kendall and sens slope tests
indices_summary <- indices_nested_ts %>%
  mutate(mk = map(data.ts, MannKendall),
         sens = map(data.ts, sens.slope))%>%
  mutate(tidy_mk = map(mk, tidy),
         tidy_sens = map(sens, tidy_sens_fun))%>%
  unnest(c(tidy_sens, tidy_mk))

indices_summary


indices_with_counts <- indices_nested_ts %>%
  mutate(data_count = as.numeric(map(.f = length,
                                     .x = data.ts)))%>%
  select(site, metric, data_count)%>%
  write_csv("indicators_data_counts.csv")
```


```{r}

#create neat data table for joining
indices_results <- indices_summary %>%
  mutate(trend = ifelse(p.value >= 0.05, 'no trend', ifelse(kendall_score > 0, "increasing", "decreasing")),
         slope =  sens.slope.estimate,
         days_per_decade = 10*abs(slope))%>%
  select(metric, trend, slope, days_per_decade, sens.p.value)

indices_results %>%
  write_csv("climate_indices_trend_results.csv")
```

```{r}
indices_for_graphs <- indices_by_year %>%
  left_join(indices_results, by = c("site", "metric"))%>%
  write_csv("indices_for_graphs.csv")


indices_for_graphs %>%
  filter(metric == "spring.24.day" | metric == "fall.24.day")%>%
  mutate(site = ifelse(site == 'summit', 'Summit', 'Pinkham'),
        value  = as.Date(value, origin = date-1),#origin is based on jan 1st being day 0, whereas doy is based on jan 1st being day 1, date-1 corrects for this by starting the origin at yearbefore-12-31
        date_value = as_date(as.numeric(value)),
        date_value = ymd(format(date_value, "1936-%m-%d")),
        metric_names = ifelse(metric == "spring.24.day", "Start", "End"))%>%
  ggplot(aes(x = date, y = date_value, shape = metric_names, color = days_per_decade))+
  geom_point(size = 1)+
  geom_line(size = .3)+
  geom_smooth(method = lm, size = .5)+
  facet_wrap(~site)+
  theme_bw()+
  labs(x = 'Year (1935-2018)', y = 'Growing Season Start and End Date', shape = "", color = "Rate of change (days per decade)")+
  scale_color_gradient(low = "red", high = "red4")+
  scale_shape_manual(values = c(16, 1), name="metric_names", guide = FALSE) +
  theme(legend.position = "bottom",
        legend.direction = "horizontal",
        legend.margin = unit(0.0, "cm"),
        legend.text = element_text(family = "Helvetica", size = 8),
        legend.title = element_text(family = "Helvetica", size = 8),
        text = element_text(family = "Helvetica", size = 10),
        axis.text = element_text( size = 10))


ggsave("freeze.24.png", width = 5.25, height = 3.28, units = "in", dpi = 500)




```


