---
title: "Monthly Temperature Trends"
author: "Annie Colgan"
date: "7/6/2020"
output: html_document
---

The purpose of this code is to calculate the mean monthly average, minimum, and maximum temperatures for Pinkham Notch and the Summit of Mount Washington and to analyze the trends in each of those metrics for each month from 1935 through 2018. Running this code produces the following:

1. 'monthly_results.csv', a csv file containing information on the mean average, minimum, and maximum temperature for each month and each site and the results of Mann Kendall and Sens Slope analysis of the trends over time. 

2. A set of three multi-panel graphs showing the trend over time in the mean average, minimum, and maximum, monthly temperature for each month. These graphs are saved as png files named 'avg_plot.png', 'max_plot.png', and 'min_plot.png', respectively. 

Required input files: "pinkham_temp.csv", "summit_temp.csv"

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Kendall)
library(tsibble)
library(lubridate)
library(timetk)
library(trend)
library(tidyquant)
library(sweep)
library(forecast)
library(broom)
library(modifiedmk)
```

**Load functions**
```{r}
#tidy() function doesn't work for sens slope. This function works instead. 
tidy_sens_fun <- function(sens){
  tidy_sens <- tibble(sens.slope.estimate = sens["estimates"] %>% unlist, 
                            sens.slope.statistic = sens['statistic'] %>% unlist,
                      sens.p.value = sens["p.value"]%>%unlist)
}
```


**1. Load data**
```{r}
#Loads Pinkham temperature data
PN_temp_df <- read_csv("pinkham_temp.csv")%>%
  mutate(date = paste(year, month, day, sep = ' ')%>% #adds date column
  ymd(),
  site = 'pinkham')%>% #adds site column
  rename(maxC = MaxT_PN, minC = MinT_PN, average = temp_c) #renames columns to match summit data
  
#Loads Summit temperature data
S_temp_df <- read_csv('summit_temp.csv')%>%
  rename(month = mo)%>% #renames month column to match pinkham data
  mutate(date = ymd(paste(year, month, day, sep = "-")), #adds date column
         site = 'summit')%>% #adds site column
  select(-jday) 

#Joins pinkham and summit data into single data frame
temp_df <- PN_temp_df %>%
  bind_rows(S_temp_df) 
```

**2. Create new data frame containing average mean, min, and max temperatures for each month**
```{r}
monthly_df1 <- temp_df %>%
  pivot_longer(cols = maxC:average, names_to = "metric", values_to = "tempC")%>%
  mutate(has.data = ifelse(is.na(tempC)== TRUE, 0, 1))%>%
  group_by(year, month, site, metric)%>%
  summarize(mean = mean(tempC, na.rm = TRUE),
            data.days = sum(has.data),
            date = first(date))#preserves a unique date for each month needed to create a ts
        
monthly_df1
```

**3. Filter based on 20 or more days of data per month threshold** Months with fewer than 20 days of data are replaced with NAs. 
```{r}
cutoff <- 20 #defines the cuttoff, can easily be changed. 
monthly_df <- monthly_df1 %>%
  mutate(mean = ifelse(data.days < cutoff, NA, mean))

monthly_df
```

**4. Calculate trend for each month x each metric using Mann Kendall and Sens Slope**
```{r}
#nest by month and metric
monthly_by_metric <- monthly_df %>%
  ungroup()%>%
  select(-year, -data.days)%>%
  filter(is.na(mean)==FALSE)%>%
  group_by(metric, month, site)%>% 
  nest()

#add column containing time series
monthly_by_metric_ts <- monthly_by_metric %>%
  mutate(data.ts = map(.x = data,
                       .f = tk_ts,
                       select = -date,
                       start = 1935,
                       freq = 12))


#create summary df with results of mann kendall and sens slope tests
monthly_metrics_summary <- monthly_by_metric_ts %>%
  mutate(mk = map(data.ts, MannKendall),
         sens = map(data.ts, sens.slope))%>%
  mutate(tidy_mk = map(mk, tidy),
         tidy_sens = map(sens, tidy_sens_fun))%>%
  unnest(c(tidy_sens, tidy_mk))

monthly_metrics_with_count <- monthly_metrics_summary %>%
  mutate(count_data = as.numeric(map(.x = data.ts,
                      .f = length)),
         percent_complete = count_data/84)%>%
  select(month, site, metric, count_data, percent_complete)%>%
  write_csv("monthly_data_counts.csv")

monthly_metrics_summary
```

**run autocorrelation check on timeseries**
```{r}
monthly_by_metric_ts %>%
 unite(name, site, month, metric)%>%
  group_by(name)%>%
  mutate(acf = map(.x = data.ts,
                   .f = acf,
                   na.action = na.pass, 
                   plot = FALSE))%>%
  mutate(plot = map(.x = acf,
                    .f = plot, 
                    main = name))

```
 
 **run partical autocorrelation check on timeseries**
```{r}
monthly_by_metric_ts %>%
  unite(name, site, month, metric)%>%
  group_by(name)%>%
  mutate(pacf = map(.x = data.ts,
                   .f = pacf,
                   na.action = na.pass, 
                   plot = FALSE))%>%
  mutate(plot = map(.x = pacf,
                    .f = plot, 
                    main = name))
```


**5. Create a neatly organized csv file containing the results of the analysis for easy reference**
```{r}
#generate results csv
monthly_results_df <- monthly_metrics_summary%>%
  mutate(mean = as.numeric(map(.x = data.ts, 
                    .f = mean,
                    na.rm = TRUE)),
         sd = as.numeric(map(.x = data.ts,
                  .f = sd,
                  na.rm = TRUE)))%>%
  mutate(trend = ifelse(p.value >= 0.1, 'no trend', ifelse(p.value < .1 & p.value >= .05, 'not significant', ifelse(kendall_score > 0, "increasing", "decreasing"))),
         slope = sens.slope.estimate,
         significant.slope = ifelse(trend == 'no trend', NA, sens.slope.estimate),
         degrees_per_decade = 10*significant.slope)%>%
  select(site, month, metric, mean, sd, trend, slope, significant.slope, degrees_per_decade, sens.p.value)

monthly_results_df1 <- monthly_results_df%>%
  pivot_wider(names_from = site, values_from = mean:sens.p.value)%>%
  relocate(mean_summit, sd_summit, trend_summit, slope_summit, significant.slope_summit, degrees_per_decade_summit, sens.p.value_summit, .after = sens.p.value_pinkham)

monthly_results_df1

monthly_results_df1 %>%
  write_csv("monthly_results.csv")
```
```{r}
#run modified mann kendall
PN_9_min <- monthly_df %>%
  filter(month == 9)%>%
  filter(site == "pinkham")%>%
  filter(metric == "minC")%>%
  filter(is.na(mean)==FALSE)%>%
  pull(mean)
  
bbsmk(PN_9_min, nsim = 2000)
mmkh(PN_9_min)

```
```{r}
PN_9_max <- monthly_df %>%
  filter(month == 9)%>%
  filter(site == "pinkham")%>%
  filter(metric == "maxC")%>%
  filter(is.na(mean)==FALSE)%>%
  pull(mean)
  
bbsmk(PN_9_max, nsim = 2000)
mmkh(PN_9_max)
```

```{r}
S_9_min <- monthly_df %>%
  filter(month == 9)%>%
  filter(site == "summit")%>%
  filter(metric == "minC")%>%
  filter(is.na(mean)==FALSE)%>%
  pull(mean)
  
bbsmk(S_9_min, nsim = 2000)
mmkh(S_9_min)
```
```{r}
S_9_max <- monthly_df %>%
  filter(month == 9)%>%
  filter(site == "summit")%>%
  filter(metric == "maxC")%>%
  filter(is.na(mean)==FALSE)%>%
  pull(mean)
  
bbsmk(S_9_max, nsim = 2000)
mmkh(S_9_max)
```


**6. Combine average, warmest, and coldest data with average, warmest, and coldest trend data**

```{r}
#create neat data table for joining
metrics_for_join <- monthly_results_df%>%
  select(month, site, metric, trend, significant.slope, degrees_per_decade)%>%
  rename(slope = significant.slope)

#join
monthly_for_graph <- monthly_df %>%
  left_join(metrics_for_join, by = c('month', 'site', 'metric'))%>%
  mutate(month_name = month(month, label = TRUE, abbr = FALSE))

```


**7. Visualize Trends**

```{r}
#Average
monthly_for_graph %>%
  filter(metric == 'average')%>%
  ggplot(aes(x = year, y = mean, color = degrees_per_decade, shape = site)) +
  geom_point()+
  geom_smooth(method = lm)+
  labs(title = "Long Term Trends in Mean Monthly Temperatures at Mount Washington and Pinkham Notch", y = "Mean Monthly Temperature (°C)", x = "Year (1935 - 2018)", color = "Degrees of Warming per Decade")+
  facet_wrap( ~ month_name, ncol = 4, scales = "free_y")+
  theme_bw()+
  theme(legend.position = "bottom")+
  scale_color_gradient(low = "red4", high = "red")

ggsave("avg_plot.png", width = 12, height = 8)#saves as plot
  

#min
monthly_for_graph %>%
  filter(metric == 'minC')%>%
  ggplot(aes(x = year, y = mean, color = degrees_per_decade, shape = site)) +
  geom_point()+
  geom_smooth(method = lm)+
  facet_wrap( ~ month_name, ncol = 4, scales = "free_y")+
  theme_bw()+
  theme(legend.position = "bottom")+
  labs(title = "Long Term Trends in Minimum Monthly Temperatures at Mount Washington and Pinkham Notch ", y = "Mean Minimum Temperature (°C)", x = "Year (1935 - 2018)", color = "Degrees of Warming per Decade")+
  scale_color_gradient(low = 'red4', high = "red")


ggsave("min_plot.png", width = 12, height = 8)

#max
monthly_for_graph %>%
  filter(metric == 'maxC')%>%
  ggplot(aes(x = year, y = mean, color = degrees_per_decade, shape = site)) +
  geom_point()+
  geom_smooth(method = lm)+
  labs(title = "Long Term Trends in Maximum MonthlyTemperatures at Mount Washington and Pinkham Notch", y = "Mean Maximum Temperature (°C)", x = "Year (1935 - 2018)", color = "Degrees of Warming per Decade")+
  facet_wrap( ~ month_name, ncol = 4, scales = "free_y")+
  theme_bw()+
  theme(legend.position = "bottom")+
  scale_color_gradient(low = "red4", high = "red")

ggsave("max_plot.png", width = 12, height = 8)

```