---
title: "Seasonal Temperature Trends"
author: "Annie Colgan"
date: "9/21/2020"
output: html_document
---
The purpose of this code is to calculate the mean annual and seasonal average temperatures for Pinkham Notch and the Summit of Mount Washington and to analyze the trends for each season from 1935 through 2018. Running this code produces the following:

1. 'seasonal_trends.csv' - a csv file containing the results of Mann Kendall and Sens Slope analysis of the trends over time for each year (annual) and each season. 
2. "seasonal_data_counts.csv" - a csv file containing information on the number of years included in the analysis for each season,
3. "seasonal_for_graph.csv" -  the csv needed to create figure 2. 

Required input files: "pinkham_temp.csv", "summit_temp.csv"

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Kendall)
library(tsibble)
library(lubridate)
library(timetk)
library(trend)
library(tidyquant)
library(sweep)
library(forecast)
library(broom)
library(tseries)
library(viridis)
library(ggpubr)
library(modifiedmk)

```
**Load functions**
```{r}
#tidy() function doesn't work for sens slope. This function works instead. 
tidy_sens_fun <- function(sens){
  tidy_sens <- tibble(sens.slope.estimate = sens["estimates"] %>% unlist, 
                            sens.slope.statistic = sens['statistic'] %>% unlist,
                      sens.p.value = sens["p.value"]%>%unlist)
}
```

**Load data**
```{r}
#Loads Pinkham temperature data
PN_temp_df <- read_csv("pinkham_temp.csv")%>%
  mutate(date = paste(year, month, day, sep = ' ')%>% #adds date column
  ymd(),
  site = 'pinkham')%>% #adds site column
  rename(maxC = MaxT_PN, minC = MinT_PN, average = temp_c) #renames columns to match summit data
  
#Loads Summit temperature data
S_temp_df <- read_csv('summit_temp.csv')%>%
  rename(month = mo)%>% #renames month column to match pinkham data
  mutate(date = ymd(paste(year, month, day, sep = "-")), #adds date column
         site = 'summit')%>% #adds site column
  select(-jday) 

#Joins pinkham and summit data into single data frame
temp_df <- PN_temp_df %>%
  bind_rows(S_temp_df) 
```


```{r}
threshold <- 20

monthly_df <- temp_df %>%
  pivot_longer(cols =  maxC:average, names_to = "metric", values_to = "tempC")%>%
  mutate(has.data = ifelse(is.na(tempC)== TRUE, 0, 1))%>%
  group_by(year, month, site, metric)%>%
  summarize(data.days = sum(has.data),
            tempC = ifelse(data.days < threshold, NA, mean(tempC, na.rm = TRUE)),
            date = first(date))
  

seasonal_df1 <- monthly_df%>%
  mutate(season = ifelse(month == 12 | month <=2, 'winter', ifelse(month <=5, 'spring', ifelse(month <= 8, 'summer', 'fall'))),
         year = ifelse(month == 12, year + 1, year))%>%#this ensures that december is averaged with the correct winter. 
  group_by(year, season, metric, site)%>%
  summarize(mean = mean(tempC),
            date = first(date))%>%
  ungroup()%>%
  filter(year > 1935 | season != 'winter') #winter 1935 is not complete because it is missing december.
  

#annual
annual_df <- monthly_df%>%
  group_by(year,metric, site)%>%
  summarize(mean = mean(tempC),
            date = first(date))%>%
  ungroup()%>%
  mutate(season = "annual")

#join annual with seasonal
seasonal_df <- seasonal_df1 %>%
  bind_rows(annual_df)


seasonal_ts <- seasonal_df %>%
  select(-year)%>%
  #filter(is.na(mean)==FALSE)%>%
  group_by(season, metric, site)%>%
  nest()%>%
  mutate(data.ts = map(.x = data, .f = tk_ts, start = 1935, select = -date))

#create a timeseries with all NAs removed for sens slope
seasonal_ts_narm <- seasonal_df %>%
  select(-year)%>%
  filter(is.na(mean)==FALSE)%>%
  group_by(season, metric, site)%>%
  nest()%>%
  mutate(data.ts.narm = map(.x = data, .f = tk_ts, start = 1935, select = -date))%>%
  rename(data.narm = data)
         
 seasonal_ts <- seasonal_ts %>%
   right_join(seasonal_ts_narm, by = c("season", "metric", "site"))
  
```
 
 
**run autocorrelation check on timeseries**
```{r}
seasonal_ts %>%
 unite(name, site, season, metric)%>%
  group_by(name)%>%
  mutate(acf = map(.x = data.ts,
                   .f = acf,
                   na.action = na.pass, 
                   plot = FALSE))%>%
  mutate(plot = map(.x = acf,
                    .f = plot, 
                    main = name))

```
 
 **run partical autocorrelation check on timeseries**
```{r}
seasonal_ts %>%
  unite(name, site, season, metric)%>%
  group_by(name)%>%
  mutate(pacf = map(.x = data.ts,
                   .f = pacf,
                   na.action = na.pass, 
                   plot = FALSE))%>%
  mutate(plot = map(.x = pacf,
                    .f = plot, 
                    main = name))
```
 
```{r}
#create summary df with results of mann kendall and sens slope tests
seasonal_trends <- seasonal_ts %>%
  mutate(mk = map(data.ts.narm, MannKendall),
         sens = map(data.ts.narm, sens.slope))%>%
  mutate(tidy_mk = map(mk, tidy),
         tidy_sens = map(sens, tidy_sens_fun))%>%
  unnest(c(tidy_sens, tidy_mk))%>%
  mutate(mean = as.numeric(map(.x = data.ts.narm,
                    .f = mean,
                    na.rm = TRUE)),
         sd = as.numeric(map(.x = data.ts.narm,
                  .f = sd,
                  na.rm = TRUE)),
         trend = ifelse(p.value >= 0.1, 'no trend', ifelse(p.value < .1 & p.value >= .05, 'not significant', ifelse(kendall_score > 0, "increasing", "decreasing"))),
         slope = sens.slope.estimate,
         significance = ifelse(trend == 'no trend' | trend == "not significant", "p > 0.05","p < 0.05"),
         degrees_per_decade = 10*slope)%>%
  select(site, season, metric, mean, sd, trend, slope, significance, degrees_per_decade, sens.p.value)

seasonal_trends %>%
  write_csv('seasonal_trends.csv')



seasonal_with_count <- seasonal_ts %>%
  mutate(data_count = as.numeric(map(.f = length,
                                     .x = data.ts.narm)),
         data_percent = data_count/84)%>%
  select(season, metric, site, data_count, data_percent)%>%
  write_csv("seasonal_data_counts.csv")
```


```{r}
PN_fall_mean <- seasonal_df %>%
  filter(season == "fall")%>%
  filter(site == "pinkham")%>%
  filter(metric == "average")%>%
  filter(is.na(mean)==FALSE)%>%
  pull(mean)
  
bbsmk(PN_fall_mean, nsim = 2000)
mmkh(PN_fall_mean)

```


```{r}
PN_annual_mean <- seasonal_df %>%
  filter(season == "annual")%>%
  filter(site == "pinkham")%>%
  filter(metric == "average")%>%
  filter(is.na(mean)==FALSE)%>%
  pull(mean)
  
bbsmk(PN_annual_mean, nsim = 2000)
mmkh(PN_annual_mean)
```

```{r}
S_annual_mean <- seasonal_df %>%
  filter(season == "annual")%>%
  filter(site == "summit")%>%
  filter(metric == "average")%>%
  filter(is.na(mean)==FALSE)%>%
  pull(mean)
  
bbsmk(S_annual_mean, nsim = 2000)
mmkh(S_annual_mean)
```

```{r}
#create neat data table for joining
seasonal_for_join <- seasonal_trends%>%
  select(site, season, metric, trend, slope, significance, degrees_per_decade)

#join
seasonal_for_graph <- seasonal_df %>%
  left_join(seasonal_for_join, by = c('season', 'site', 'metric'))%>%
  write_csv("seasonal_for_graph.csv")
  
```

