---
title: "Snow Trends"
author: "Annie Colgan"
date: "9/29/2020"
output: html_document
---

This script was written to calculate and analyze the trends in the following metrics: first snowfall, last snowfall, snow season start and end date, total snow, and maximum snow depth for the Pinkham Notch and Summit snow records. Running this code produces the following:

1. "snow_results.csv" - a csv file containing results of trend analysis for each metric x each site including slopes and p values. 
2. "snow_counts.csv" - a csv file countaining information on the number of years included in the analysis for each metric. 
3. "snow_for_graphs.csv" - a csv file with all information needed to create figure 3. 
4. "snow_seaon_results.csv" - csv file with results of trend analysis of the snow season start and end date data from Seidel et al. 2009. 

Required input files: "PN_snow.csv", "summit_snow.csv",  "Summit_SE_continous_snow.csv", "Pinkham_SE_continous_snow.csv"



The following lines of code call the packages used in this script. You may need to download some of these packages if you don't already have them using the 'install.packages' function.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(Kendall)
library(timetk)
library(trend)
library(tidyquant)
library(sweep)
library(forecast)
library(broom)
library(slider)
library(BBmisc)#which.first

```

** Define Functions: **
The following code defines key functions that are used elsewhere in the script. 
```{r}
### WINTER JULIAN DAY FUNCTION ###
#converts a date into a winter day of year, accounts for leap years
winter_doy_fun <- function(date){
  if(is.na(date == TRUE)){
    return(NA)
  }
  else{
  year <- year(date)
  month <- month(date)
  jday <- yday(date)
  winter.jday <- ifelse(leap_year(year) == TRUE, ifelse(month <=6, jday + 183, jday - 182), ifelse(month <= 6, jday + 184, jday -181))
}}

### UNDO WINTER JULIAN DAY FUNCTION ###
#converts a wdoy back into a date
#careful - this produces a list object
undo_winter_doy_fun <- function(wdoy, winter.year){
  if(is.na(wdoy == TRUE)){
    return(NA)
  }
  else{
    year <- ifelse(wdoy <= 184, winter.year-1, winter.year)
    doy <- ifelse(leap_year(year) == FALSE, ifelse(wdoy <= 184, wdoy + 181, wdoy - 184), ifelse(wdoy <= 185, wdoy + 181, wdoy -184))
    origin.date <- lubridate::ymd(year, truncated = 2)
    date <- as.Date(doy, origin = as.character(origin.date - 1))
    return(date)
  }
}


### Snow Season Function ###
#This function calculates the length of the snow season as defined by the longest period with continuous snow depth greater than 2.54cm with no thaw period greater than 4 days (as in Seidel et al. 2009). 

#input: a dataframe containing a date column and a snow.depth column (mm) with data for a single winter year (July 1st through June 30th). 

#output: a two element list called "snow.season" containing "start.date" (the date corresponding to the start of the snow season) and "end.date" (the date corresponding to the end of the snow season).


snow_season_length_fun <- function(data){
  #create time series
  snow.date <- data$date
  snow.ts1 <- zoo(data$snow.depth.filled, snow.date)
  snow.ts<- snow.ts1[!is.na(snow.ts1)]
  
  #initialize variables
  count <- 0 
  thaw.period <- 0
  high.count <- 0
  start.date <- NA
  running.start.date <- NA
  end.date <- NA
  index <- 0
  
  #looping through days
  for( i in snow.ts) {
    index <- index + 1
    if (i >= 25.4){
      count = count + 1
      thaw.period = 0
      if(count == 1){running.start.date <- snow.date[index]}
      }
    else if (i <25.4){
      thaw.period = thaw.period + 1
      if (thaw.period > 4){
        if(count > high.count) {
          end.date <- snow.date[index-5]
          high.count <- count
          start.date <- running.start.date
          }
        count = 0}
      else if (thaw.period <= 4){
        count = count + 1}}
  }
  snow.season <- data.frame(start.date, end.date)
  
  return(snow.season)}



### TIDY SENS SLOPE FUNCTION ###
#tidy() function doesn't work for sens slope. This function works instead. 
#input: the output of the sens function
#output: the sens estimate, sens statistic, and sens.p.value, unlisted
tidy_sens_fun <- function(sens){
  tidy_sens <- tibble(sens.slope.estimate = sens["estimates"] %>% unlist, 
                            sens.slope.statistic = sens['statistic'] %>% unlist,
                      sens.p.value = sens["p.value"]%>%unlist)
}



```


**1. Load snow data and join sites: **
The following code loads the data from csv files and joins them into a single table. 
```{r}
###SNOW DATA###

#Load Pinkham snow data
#starts Jan 1, 1930
PN_snow_df <- read_csv('PN_snow.csv')%>%
  select(DATE, `SNOW(MM)`, `SNAW (MM)`)%>%
  rename(date = DATE, snow.fall = `SNOW(MM)`, snow.depth = `SNAW (MM)`)%>%#renames columns
  mutate(date = mdy(date),#converts date to type date
         site = 'pinkham',
         depth.missing = ifelse(is.na(snow.depth)==TRUE, 1, 0))%>%
  mutate(snow.depth.filled = snow.depth)%>%
  fill(snow.depth.filled)#fills missing snow depth value with previous, complete data is required to run snow season function


#Load Summit snow data
#starts Jan 1, 1948
S_snow_df <- read_csv('summit_snow.csv')%>%
  mutate(date = mdy(Date),#adds date column with data in type date
         Snowfall = as.numeric(ifelse(Snowfall == "M", NA, ifelse(Snowfall == "T", (0), Snowfall))), #replaces M with NA and T with 0, to be consistent with PN
         SnowDepth = as.numeric(ifelse(SnowDepth == "M", NA, ifelse(SnowDepth == "T", (.5), Snowfall))),#replaces T with .5, half the 1" threshold for snow depth
         snow.fall = Snowfall*25.4, #converts to mm
         snow.depth = SnowDepth*25.4, #converts to mm
         site = 'summit',
         depth.missing = ifelse(is.na(snow.depth) == TRUE, 1, 0))%>% 
  select(date, snow.depth, snow.fall, site, depth.missing)%>%
  mutate(snow.depth.filled = snow.depth)%>%
  fill(snow.depth.filled)
  #fills missing snow depth value with previous, complete data is required to run snow season function


###JOIN DATA###
snow_df <- PN_snow_df %>%
  bind_rows(S_snow_df)

snow_df
```


**2 Add columns for winter year, snow fall presence absence (1/0), and missind data presence absence to dataframe**

```{r}
#add winter.year column, snow.day column (1 if snow.fall occured, 0 if no snow.fall), and columns for calculating number of missing snowfall days. 
snow_df2 <- snow_df%>%
  mutate(year = year(date),
         month = month(date),
         winter.jday = winter_doy_fun(date), #adds winter julian day column
        winter.year = ifelse(month <= 6, year, year+1), #add winter.year column corresponding to the year for Jan-Jun
         snow.day = ifelse(snow.depth > 0, winter.jday, NA),
         fall.missing = ifelse(is.na(snow.fall) == TRUE, 1, 0))
          #information on missing snow depth data was previously added

snow_df2
      
```

**3. Deal with missing data calculate metrics**
```{r}

### DEFINE THRESHOLDS FOR DEALING WITH MISSING DATA ###
snow.threshold <- 10
snow.season.threshold <- 30
winter.start.pn <- 123 #November 1st as wdoy
winter.end.pn <- 335 #May 31 as wdoy

#Adjust missing data to only include relevant months (Oct.- May)
snow_adjusted <- snow_df2 %>% 
  mutate(depth.missing =  ifelse(winter.jday > winter.start.pn & winter.jday < winter.end.pn & depth.missing == 1, 1, 0), #adjusts depth missing to only care about days between october 1st and may 31 
         fall.missing = ifelse(winter.jday > winter.start.pn & winter.jday < winter.end.pn & fall.missing == 1, 1, 0))
  
### CALCULATE METRICS ###
#calculate metrics that require summarizing
metrics_by_year <- snow_adjusted%>%
  group_by(site, winter.year)%>%
  summarize(depth.missing.sum = sum(depth.missing),
            first.snow = ifelse(sum(depth.missing) < snow.threshold, 
                                min(snow.day, na.rm = TRUE), NA),
            last.snow = ifelse(sum(depth.missing) < snow.threshold, 
                               max(snow.day, na.rm = TRUE), NA),
            tot.snow = ifelse(sum(fall.missing) < snow.threshold, sum(snow.fall, na.rm = TRUE), NA),
            max.snow.fall = ifelse(sum(fall.missing)< snow.threshold, max(snow.fall, na.rm = TRUE), NA),
            max.snow.depth = ifelse(sum(depth.missing) < snow.threshold, max(snow.depth, na.rm = TRUE), NA)
  )
  
#calculate metrics that require nesting and join with other metrics
snow_season <- snow_adjusted %>%
  filter(winter.year > 1930 & winter.year <= 2018) %>%
  select(site, winter.year, snow.depth.filled, date) %>%
  group_by(site, winter.year)%>%
  nest()%>%
  left_join(metrics_by_year, by = c("site", "winter.year"))%>%
  mutate(snow.season  = map(.x = data,
                             .f = snow_season_length_fun))%>%
  unnest(snow.season)%>%
 mutate(snow.season.start = ifelse(depth.missing.sum < snow.season.threshold, as.numeric(map(.x = start.date,
                                  .f = winter_doy_fun)), NA),
          snow.season.end = ifelse(depth.missing.sum < snow.season.threshold, as.numeric(map(.x = end.date,
                                .f = winter_doy_fun)), NA))%>%
  filter(!( site == "summit" & winter.year == 1948))

snow_metrics <- snow_season %>%
  select(site, winter.year, first.snow, last.snow, snow.season.start, snow.season.end)%>%
  write_csv("snow_for_comparison.csv")
         
```


**4. Calculate Sens Slope and Mann Kendall**
```{r}
#reorganize data into longer format for more efficient processing.
snow_longer <- snow_season %>%
  ungroup()%>%
  mutate(date = ymd(winter.year, truncated = 2))%>%
  select(-data,-depth.missing.sum, -start.date, -end.date, -winter.year)%>%
  pivot_longer(cols = first.snow:snow.season.end, names_to = 'metric', values_to = 'value')
  
#nest data by metric and site
snow_nested <- snow_longer %>% 
  filter(is.na(value)==FALSE)%>%
  mutate(value = as.numeric(value))%>%
  group_by(metric, site)%>% 
  nest()

#add column containing time series
snow_nested_ts <- snow_nested %>%
  mutate(data.ts = map(.x = data,
                       .f = tk_ts,
                       select = -date))

#create summary df with results of mann kendall and sens slope tests
snow_summary <- snow_nested_ts %>%
  mutate(mk = map(data.ts, MannKendall),
         sens = map(data.ts, sens.slope))%>%
  mutate(tidy_mk = map(mk, tidy),
         tidy_sens = map(sens, tidy_sens_fun))%>%
  unnest(c(tidy_sens, tidy_mk))

snow_summary

```

**run autocorrelation check on timeseries**
```{r}
snow_nested_ts %>%
 unite(name, site, metric)%>%
  group_by(name)%>%
  mutate(acf = map(.x = data.ts,
                   .f = acf,
                   na.action = na.pass, 
                   plot = FALSE))%>%
  mutate(plot = map(.x = acf,
                    .f = plot, 
                    main = name))

```
 
 **run partical autocorrelation check on timeseries**
```{r}
snow_nested_ts %>%
  unite(name, site, metric)%>%
  group_by(name)%>%
  mutate(pacf = map(.x = data.ts,
                   .f = pacf,
                   na.action = na.pass, 
                   plot = FALSE))%>%
  mutate(plot = map(.x = pacf,
                    .f = plot, 
                    main = name))
```

**5. Create neat csv with results and df for graphing **
```{r}
#create neat data table for joining
snow_for_join <- snow_summary %>%
  mutate(trend = ifelse(p.value >= 0.05, 'no trend', ifelse(kendall_score > 0, "increasing", "decreasing")),
         slope = sens.slope.estimate,
         days_per_decade = 10*abs(slope))%>%
  select(metric, trend, slope, days_per_decade, sens.p.value)

snow_for_join %>% write_csv("snow_results.csv")

snow_with_count <- snow_nested_ts %>%
  mutate(data_count = as.numeric(map(.f = length,
                                     .x = data.ts)))%>%
  select(metric, site, data_count)%>%
  write_csv("snow_counts.csv")

#join
snow_for_graphs <- snow_longer %>%
  left_join(snow_for_join, by = c('metric', 'site'))%>%
  filter(metric == "tot.snow"| metric == "max.snow.depth"| metric == "max.snow.fall")%>%
  write_csv("snow_for_graphs.csv")

```

Snow season trend analysis using snow season start and end from Seidel et al. 

```{r}
#load data
summit_snowseason <- read_csv("Summit_SE_continous_snow.csv")%>%
  mutate(site = "Summit")
pinkham_snowseason <- read_csv("Pinkham_SE_continous_snow.csv")%>%
  mutate(site = "Pinkham",
         JdayStart = as.numeric(JdayStart),
         JdayEnd = as.numeric(JdayEnd))

#wrangle data into nested form to easily run trend analysis
snow_season <- summit_snowseason %>%
  bind_rows(pinkham_snowseason)%>%
  mutate(Date = ymd(Year, truncated = 2))%>% #adds date column needed for generation of timeseries type
  select(-Year)%>%
  pivot_longer(cols = JdayStart:JdayEnd, names_to = "metric")%>%
  filter(is.na(value)==FALSE)%>%
  group_by(site, metric)%>%
  nest()

#add column containing time series
snow_season_ts <- snow_season %>%
  mutate(data.ts = map(.x = data,
                       .f = tk_ts,
                       select = -Date))


#create summary df with results of mann kendall and sens slope tests
snow_season_summary <- snow_season_ts %>%
  mutate(mk = map(data.ts, MannKendall),
         sens = map(data.ts, sens.slope))%>%
  mutate(tidy_mk = map(mk, tidy),
         tidy_sens = map(sens, tidy_sens_fun))%>%
  unnest(c(tidy_sens, tidy_mk))

snow_season_results <- snow_season_summary %>%
  select(site, metric, sens.slope.estimate, sens.p.value)%>%
  write_csv("snow_season_results.csv")

```

